Properties and Bayesian fitting of restricted Boltzmann machines

### 2018-09-30 writing/paper-rbm/paper.tex
revisions

\caption{\label{fig:uninterp}The absolute difference between coordinates of E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ expectations given independence for an RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as$\theta_{11}$, the dependence parameter, deviates from zero.}\label{fig:uninterp}
\end{figure}

@@

The \(-1/1\) data encoding has the benefit of providing a
guaranteed-to-be non-degenerate model at
\(\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}\),
where the zero vector then serves as the natural center of the parameter
space and induces the simplest possible model properties for the RBM
(i.e., at \(\boldsymbol \theta = \boldsymbol 0\), all variables are
independent uniformly distributed on \(\{-1,1\}^\nv\)). The proof of
this and further exploration of the equivalence of the
\(\boldsymbol \theta\) parameterization of the RBM model class and
parameterization by \(\boldsymbol \mu(\boldsymbol \theta)\) is in the
on-line supplementary materials. Hence, while from some computing
perspectives \(0/1\) coding might seem most natural, the \(-1/1\) coding
is far more convenient and interpretable from the point of view of
statistical modeling, where it makes parameters simply interpreted in
terms of symmetrically defined main effects and interactions. Under the
data encoding \(-1/1\), the parameter space centered at 0 is also
helpful for framing parameter configurations that are undesirably large
(i.e., these are naturally parameters that have moved too far away from
\(\bm{0}\) where the RBM model is anchored to be trivially describable
and completely problem-free). In light of all of these matters we will
henceforth employ the \(-1/1\) coding.

@@

For simplicity, consider a model with two visible variables
\((V_1,V_2)\) and one hidden \(H_1\) so that there are four possible
data realizations for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\), and we
may express the model probabilities as \[
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
\] for \((v_1,v_2)\in \{-1, 1\}^2\), in terms of real-valued parameters
\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
Given some specified cell probabilities, say \begin{equation}
\label{eq:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} for the outcomes \((\pm 1, \pm 1)\), the online appendix
provides technical details to show parameter values
\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
may be chosen to match any given cell probabilities with arbitrary
closeness. In fact, when the cell probabilities \eqref{eq:0} are all
strictly positive, parameters in the RBM model can be specified to
reflect these probabilities exactly. And, when one or more of the cell
probabilities \eqref{eq:0} are zero, the corresponding RBM probabilities
\(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
may never be identically zero (due to exponential terms in the model)
but parameters still can be selected to make the appropriate RBM cell
probabilities arbitrarily small. Furthermore, as development in the
online appendix shows, not only is it possible to approximate any
distribution on the visibles arbitrarily well (cf. Montufar and Ay
2011), but dramatically different parameter settings can induce the same
RBM model (beyond mere symmetries in parameterization). A further (and
somewhat odd) consequence of the RBM parameterization is then that when
fitting the RBM model by likelihood-based methods, we already may know
the nature of the answer before we begin: namely, such fitting simply
will reproduce the empirical distribution of the training data if
sufficiently many hiddens are in the model. There then can be no model
refinements or smoothing from the RBM. That is, based on a random sample
of vectors of visible variables, the model for the cell probabilities
with the highest likelihood over \emph{all possible model classes}
(i.e., RBM-based or not) is the empirical distribution, and the
over-parameterization of the RBM model itself ensures that this
empirical distribution can be arbitrarily well-approximated.

For illustration, considering the simple example from above with \(n\)
iid observations, each consisting of two realized visibles \((V_1,V_2)\)
the empirical cell frequencies from the sample are the highest
likelihood choices for the cell probabilities
\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in \eqref{eq:0} before
any parametric model is introduced for improved analysis, and the
discussion above indicates that RBM model parameters can be chosen to
re-create this empirical distribution to an arbitrarily close degree.
Not only does RBM model fitting based on ML seek to reproduce the
empirical distribution, whenever this empirical distribution contains
empty cells, fitting steps for the RBM model further will aim to choose
parameters that necessarily diverge to infinity in magnitude in order to
zero-out the corresponding RBM cell probabilities. In data applications
with a large sample space, it is unlikely that the training set will
include at least one of each possible vector outcome (unlike this small
example). This implies that some RBM model parameters must diverge to
\(+\infty\) to mimic the empirical distribution with empty cells and, as
we have already discussed in Section
\ref{explorations-of-model-properties-through-simulation}, large
magnitudes of \(\boldsymbol \theta\) lead to model impropriety in the
RBM.

@@

Ultimately, it is not unquestionably clear that RBM models are useful as
generative models. As a concern, without appropriate generative behavior
in an RBM model, the uncertainty in estimated model parameters becomes
impossible to assess realistically and further, the model can lose
useful application to prediction problems (e.g., realistic predictive
distributions). In the case of classification, predictive distributions
may not be the ultimate goal, but when S-instability is present, small
(imperceptible) differences in the data may lead to greatly different
probabilities and thus greatly different classifications. For these
reasons, we are skeptical about RBMs as probabilistic conceptualizations
for data generation.

@@

\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``On the
Propriety of Restricted Boltzmann Machines.'' \emph{Unpublished}.

>>>>

\caption{\label{fig:uninterp}The absolute difference between coordinates of E$\left[\boldsymbol X | \boldsymbol \theta\right]$ model expectations and E$\left[\boldsymbol X | \boldsymbol \theta^* \right ]$ expectations given independence for an RBM with one visible and one hidden node. As an indicator of uninterpretability, note that differences in expectations increase as $\theta_{11}$, the dependence parameter, deviates from zero.}\label{fig:uninterp}
\end{figure}

@@

The \(-1/1\) data encoding has the benefit of providing a
guaranteed-to-be non-degenerate model at
\(\boldsymbol \theta = \boldsymbol 0 \in \mathbb{R}^{\nh + \nv + \nh*\nv}\),
where the zero vector then serves as the natural center of the parameter
space and induces the simplest possible model properties for the RBM
(i.e., at \(\boldsymbol \theta = \boldsymbol 0\), all variables are
independent uniformly distributed on \(\{-1,1\}^\nv\)). The proof of
this and further relations between the \(-1/1\) data encoding and the
\(\boldsymbol \theta\) parameterization of the RBM model class are given
in the on-line supplementary materials. Based on this material,
Proposition \ref{prp:three} also briefly summarizes how the \(-1/1\)
data encoding induces some attractive theoretical properties in
characterizing RBM models.

\begin{proposition}
\label{prp:three}
Let $\nv, \nh \geq 1$ and suppose the random vector $X =(V_1,...,V_{\nv},H_1,...,H_{\nh})$ assumes values in $\{-1,1\}^{\nv+\nh}$ and follows the RBM$(\nv, \nh, \boldsymbol \theta)$ distribution \eqref{eq:pmf} with parameter $\boldsymbol \theta\in \mathbb{R}^{\nv+\nh+\nv*\nh}$. Let $\bm{\theta}_1,\bm{\theta}_2 \in \mathbb{R}^{\nv+\nh+\nv*\nh}$ and let $P_i$ and $\E_i$ denote the full joint probability and expectation under $\bm{\theta}_i$, $i=1,2$. Then, the following are equivalent
\begin{enumerate}
\setlength{\itemsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
\item $\E_1 V_i=\E_2 V_i$, $\E_1 H_j=\E_2 H_j$ and $\E_1 V_i H_j =\E_2 V_i H_j$ for all $i=1,\ldots, \nv$, $j=1,\ldots,\nh$.
\item $\bm{\theta}_1 =\bm{\theta}_2$.
\item $P_1$ and  $P_2$ are the same distribution for $\boldsymbol X$.
\end{enumerate}
Additionally, the following three statements are equivalent
\begin{enumerate}
\setlength{\itemsep}{0pt}
\setlength{\parskip}{0pt}
\setlength{\parsep}{0pt}
\item $\E V_i=0$, $\E H_j=0$ and $\E V_i H_j =0$ for all $i=1,\ldots,\nv$, $j=1,\ldots,\nh$.
\item $\bm{\theta} = \bm{0}\in \mathbb{R}^{\nv+\nh+\nv*\nh}$.
\item $V_1,\ldots,V_\nv, H_1,\ldots,H_\nh$ are independent random variables assuming $-1/1$ with equal probability $1/2$.
\end{enumerate}
\end{proposition}

Hence, while from some computing perspectives \(0/1\) coding might seem
most natural, the \(-1/1\) coding is far more convenient and
interpretable from the point of view of statistical modeling, where it
makes parameters simply interpreted in terms of symmetrically defined
main effects and interactions. Under the data encoding \(-1/1\), the
parameter space centered at 0 is also helpful for framing parameter
configurations that are undesirably large (i.e., these are naturally
parameters that have moved too far away from \(\bm{0}\) where the RBM
model is anchored to be trivially describable and completely
problem-free). In light of all of these matters we will henceforth
employ the \(-1/1\) coding.

@@

For simplicity, consider a model with two visible variables
\((V_1,V_2)\) and one hidden \(H_1\) so that there are four possible
data realizations for \((V_1,V_2)\) given by \((\pm 1, \pm 1)\), and we
may express the marginalized model probabilities as \[
P(V_1=v_1,V_2=v_2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}) \propto  \exp\left( v_1  \theta_{v_1} + v_2 \theta_{v_2}\right) \sum_{h \in \{\pm 1\}}\exp\left( h[ \theta_{h_1}  + \theta_{11} v_1 + \theta_{21} v_2] \right),
\] for \((v_1,v_2)\in \{-1, 1\}^2\), in terms of real-valued parameters
\(\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21}\).
Given some specified cell probabilities, say \begin{equation}
\label{eq:0}
0\leq p_{(-1,-1)},\;p_{(1,-1)}, \;p_{(-1,1)},\;p_{(1,1)},
\end{equation} for the outcomes \((\pm 1, \pm 1)\), the online appendix
provides technical details to show parameter values
\((\theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11}, \theta_{21})\)
may be chosen to match any given cell probabilities with arbitrary
closeness. In fact, when the cell probabilities \eqref{eq:0} are all
strictly positive, parameters in the RBM model can be specified to
reflect these probabilities exactly. And, when one or more of the cell
probabilities \eqref{eq:0} are zero, the corresponding marginalized RBM
probabilities
\(P(V_1=v_1,V_2=_v2| \theta_{v_1}, \theta_{v_2}, \theta_{h_1}, \theta_{11},\theta_{21})\)
may never be identically zero (due to exponential terms in the model)
but parameters still can be selected to make the appropriate RBM cell
probabilities arbitrarily small. Furthermore, as development in the
online appendix shows, not only is it possible to approximate any
distribution on the visibles arbitrarily well (cf. Montufar and Ay
2011), but dramatically different parameter settings can induce the same
RBM model (beyond mere symmetries in parameterization). A further (and
somewhat odd) consequence of the RBM parameterization is then that when
fitting the RBM model by likelihood-based methods, we already may know
the nature of the answer before we begin: namely, such fitting simply
will reproduce the empirical distribution of the training data if
sufficiently many hiddens are in the model. There then can be no model
refinements or smoothing from the RBM. That is, based on a random sample
of vectors of visible variables, the model for the cell probabilities
with the highest likelihood over \emph{all possible model classes}
(i.e., RBM-based or not) is the empirical distribution, and the
over-parameterization of the RBM model itself ensures that this
empirical distribution can be arbitrarily well-approximated.

For illustration, considering the simple example from above with \(n\)
iid observations, each consisting of two realized visibles \((V_1,V_2)\)
the empirical cell frequencies from the sample are the highest
likelihood choices for the cell probabilities
\(p_{(-1,-1)},p_{(1,-1)}, p_{(-1,1)},p_{(1,1)}\) in \eqref{eq:0} before
any parametric model is introduced for improved analysis, and the
discussion above indicates that RBM model parameters can be chosen to
re-create this empirical distribution to an arbitrarily close degree.
Not only does RBM model fitting based on ML seek to reproduce the
empirical distribution, whenever this empirical distribution contains
empty cells, fitting steps for the RBM model further will aim to choose
parameters that necessarily diverge to infinity in magnitude in order to
zero-out the corresponding marginalized RBM cell probabilities. In data
applications with a large sample space, it is unlikely that the training
set will include at least one of each possible vector outcome (unlike
this small example). This implies that some RBM model parameters must
diverge to \(+\infty\) to mimic the empirical distribution with empty
cells and, as we have already discussed in Section
\ref{explorations-of-model-properties-through-simulation}, large
magnitudes of \(\boldsymbol \theta\) lead to model impropriety in the
RBM.

@@

While interpretability of parameters may not always be of primary
importance in application of an RBM model (e.g.~in some applications to
prediction/classification), we find that the issues of stability and
degeneracy go hand in hand with those of interpretability, and the
former should not be ignored. When S-instability is present, small
(imperceptible) differences in training data and corresponding fitted
models may lead to greatly different probabilities (and thus
substantially different classifications). Ultimately, it is not clear
that RBM models are typically useful as generative models for
observables. Without appropriate generative behavior in an RBM model,
uncertainty in estimated model parameters becomes impossible to
realistically assess. Further, the model cannot support other inferences
important in many prediction problems (e.g., realistic predictive
distributions). For these reasons, we are skeptical about RBMs as widely
useful probabilistic constructs, at least in their present incarnation.

@@

\leavevmode\hypertarget{ref-kaplan2016propriety}{}%
Kaplan, Andee, Daniel Nordman, and Stephen Vardeman. 2017. ``On the
Instability and Degeneracy of Deep Learning Models.'' \emph{arXiv
Preprint arXiv:1612.01159}.
